Tool-driven Revolutions in Archaeological Science

### 2018-09-09 analysis/paper/jcaa-paper.Rmd
begin of elaborate discussion

The use of R is on the rise among archaeologists using natural and physical science methods, and has concrete advantages for authors that publish papers that cite R as they tend to be referenced more often. This also is an indicator for the increase in popularity of R.

The use of R or any other scripting languages enables a paradigm-shift, as it offers the possibility to share one's code. The sharing of code has been identified in other disciplines as an amplifier of methodological change [@touchon_and_mccoy_2016]. It also enables researchers to not just reproduce the work of others, but to also evaluate them according to the scientific method. Thus, the sharing of code may be at the root of a tool-driven paradigm-change as desribed in the work of @Galison_1997.

Speaking with Kuhn's view on paradigm-change, a change in the view point of researchers is necessary. We propose that a change of point of view of the archaeological researcher is needed: She or he, who works with computational methods, needs to view him/herself as a scientist, who needs to adhere to the scientific method, i. e. making the research reproducible for others.

>>>>

In light of the discussion in the philosophy of science, whether paradigm-shifts are tool- or idea-driven, we offer the hypothesis that a tool-driven change of paradigms is underway in the biological sciences and, more subtly, in archaeology. The tool which drives this change is the use of scripting languages in computational approaches of these disciplines. We could show that the use of R is on the rise among archaeologists using natural and physical science methods, and has concrete advantages for authors that publish papers that cite R as they tend to be referenced more often. This also is an indicator for the increase in popularity of R.

The use of R or any other scripting languages enables a paradigm-shift, as it offers the possibility to share one's code. The sharing of code has been identified in other disciplines as an amplifier of methodological change [@touchon_and_mccoy_2016]. It also enables researchers to not just reproduce the work of others, but to also evaluate them according to the scientific method. Thus, the sharing of code may be at the root of a tool-driven paradigm-change as desribed in the work of @Galison [-@Galison_1997]. This change does not take place revolutionary-quick and does not replace older ideas as such (as has been a critique of Kuhn's idea-driven concept of paradigm shift by others [@meltzer_1979, @trigger_2006]). Rather, this enables the archaologist to return to the tenet of the scientific method which demands the reproducibility of results. We do not want to claim archaeologists did not want to create reproducible research before the advance of scientific scripting languages. There is a longstanding tradition of publishing catalogues, lists and maps showing the data used for analyses. In the digital age, though, many archaeological tasks have shifted from the descriptive and qualitative aspects to quantitative and computational analyses. This shift also mandates a change in documentation of analyses to enable other researchers to reproduce published results. Modern archaeology has often been lacking in this aspect, as the research has been undertaken using point-and-click software, which complicates the reproducibility of anlayses. As we could show, the advance of scripting languages takes place mostly in the part of archaeology that is in close contact with biology (zooarchaeology) or focuses on scientific analyses.

Speaking with Kuhn's view on paradigm-change, a change in the view point of researchers seems to be happening in these areas: She or he, who works with computational methods, needs to view him/herself as a scientist, who needs to adhere to the scientific method, i. e. making the research reproducible for others.
